A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-09-01 analysis/paper/paper.Rmd
Minor edits after read-through

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author: 
- 'Isak Roalkvam'
- 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - Isak Roalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [UiO]
  #     correspondence: true
# institute:
#   - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating which is widely applied. However, while the potentially immense benefits of an additional source of temporal data separate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more ubiquitous dating techniques in archaeology. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. This is done by means of Monte Carlo simulation, which is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement. The findings largely confirm previous evaluations of this relationship, indicating that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.

@@

The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g. @akerlund1996; @bjerck2005; @gjerde2021; @loken1977; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009; @wikell2009].

The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was established by the end of the 19th century [@degeer1896], and was first applied as a dating method at the turn of the century [@brogger1905; @hollender1901]. Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g. @berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @olsen1984; @nummedal1923; @mikkelsen1975; @mjærum2022; @shetelig1922; @solheim2018]. The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates. However, given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

Despite its important role for Norwegian Stone Age archaeology, the applicability of dating by reference to shoreline displacement has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age, and in turn have this inform an improved method for shoreline dating. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shoreline is inherently spatial. Taking inspiration from studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g. @bevan2013a; @crema2012; @crema2010; @crema2015; @yubero-gomez2016], a similar approach is adopted here.

@@

Relative sea-level (RSL) can be defined as the mean elevation of the surface of the sea relative to land, or, more formally, the difference in elevation between the geoid and the surface of the Earth as measured from the Earth's centre [@shennan2015]. Variation in this relative distance follow from a range of effects [e.g. @milne2009]. Of central importance here is eustasy and istostasy. The eustatic sea-level is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015]. The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans [@mörner1976]. Istostasy, on the other hand, pertains to adjustments in the crust to regain gravitational equilibrium relative to the underlying viscous mantle. This can be the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed. These effects thus causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight [@milne2015].

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g. @hughes2016; @stroeven2016, see Figure \@ref(fig:deglaciation)], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g. @mörner1979; @svendsen1987]. In other words, the RSL has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable RSL or been subject to marine transgression [e.g. @romundset2015; @svendsen1987]. These conditions are directly reflected in the archaeological record. In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time-spans, creating a mix of settlement phases that are difficult to disentangle [e.g. @hagen1963; @reitan2009]. Transgression phases, on the other hand, can lead to complete destruction of the sites, bury them in marine sediments, or in the outermost periphery, leave them still submerged today [@bjerck2008; @glorstad2020]. This can lead to a hiatus in the archaeological record for certain sub-phases in the impacted areas. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 220m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @romundset2018; @sørensen1979]. This makes the region especially useful for evaluating the assumption of a shore-bound settlement pattern over a long and continuous time-span.

```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from c. 17--8 kyr BCE. The study area defined later in the text is marked with a red outline (deglaciation data from Hughes et al. 2016, but see also Romundset et al. 2019 in relation to the study area).", out.width = "525px", fig.align = "center", regfloat = TRUE}

@@

The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than an earliest possible date for when a site could have been in use [see review by @nordqvist1999]. The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @glorstad2002; @glorstad2003; @glorstad2004; @jaksland2001; @jaksland2012; @jaksland2012a; @jaksland2014a; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below]. Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time. This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @mjærum2022; @nielsen2021b; see also @jorgensen2020; @tallavaara2020]. In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963; @johansen1997].

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@brogger1905, 166; @bjerck1990; @bjerck2017; also discussed by @berg-hansen2009; @bergsvik2009]. The same logic has also been extended to the hinter- and inland regions, where sites are to be predominantly located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@prescott2020; cf. @solheim2021]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm is to have led site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2020; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2012]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227]. @nielsen2021b has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple foraging/agricultural dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g. @amundsen2006; @solheim2012, 74; @ostmo1988]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line with the transition to the Early Neolithic around 3900 BCE, and finally a decisive shift with the Late Neolithic c. 2400 BCE.

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example reflected in surveys, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g. @berg-hansen2009; @eskeland2017; @nummedal1923]. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date-span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was not shore bound [e.g. @jaksland2014; @johansen1963; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @ilves2011].

One of the more extensive evaluations of the relationship between archaeological radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites. However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates. This is based on typological and technological characteristics of the assemblages. Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these dates are entirely erroneous can be difficult to evaluate [e.g. @persson2008; @schulke2020]. However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas as evidenced by artefact inventories or multiple site features. Not remnants of stays as ephemeral to only be discernible by isolated features or dubious ^14^C-dates. The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent on, and follow the discretion of the original excavation reports.

Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @akerlund1995; @åstveit2018; @solheim2020; see also @bjerck2008a; @kleppe1985; @ramstad2009]. This approach has a couple of limitations. First of all, the displacement curves are sometimes applied directly to larger study areas, analogous to what @borreggine2022 term a bathtub model, with only some studies having taken the variable uplift-rates into account when performing this comparison [e.g. @åstveit2018; @fossum2020; @møller1987; @persson2008]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and help move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.

# Data
To get at the relationship between sites and the contemporaneous shoreline, this analysis was dependent on a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west. This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].

The employed shoreline displacement data is based on the so-called isolation basin method [e.g. @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments. Each basin thus represent a high precision sea-level index point (SLIP) which are combined using what has been termed the isobase method to devise a continuous time series for RSL-change adjusted to a common isobase. To minimise the impact of variable uplift rates, the cored basins are therefore located in a as constrained area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more severe towards the north-east, meaning that this needs to be adjusted for in the case that any basins are located any significant distance from the common isobase perpendicular to this gradient (Figure \@ref(fig:map-iso)). The SLIPs indicate the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves, based on the present day tidal range. This is assumed to have been the same throughout the Holocene [@sørensen2014, 44]. The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021 30cm at the standard port Helgeroa in Larvik]. Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but also hold potential error as related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g. @romundset2011; @romundset2019; for alternative approaches see e.g. @barnett2020; @cahill2016; @creel2022]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

>>>>

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author: 
- 'Isak Roalkvam'
- 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - Isak Roalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [UiO]
  #     correspondence: true
# institute:
#   - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating which is widely applied. However, while the potentially immense benefits of an additional source of temporal data separate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more established dating techniques in archaeology. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. Monte Carlo simulation is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement. The findings largely confirm previous hypotheses that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.

@@

The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process created throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g. @akerlund1996; @bjerck2005; @gjerde2021; @loken1977; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009; @wikell2009].

The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was proposed at the end of the 19th century [@degeer1896], and it was first applied as a dating method at the turn of the century [@brogger1905; @hollender1901]. Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g. @berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @olsen1984; @nummedal1923; @mikkelsen1975; @mjærum2022; @shetelig1922; @solheim2018]. The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates. However, given the coarse and fuzzy resolution of established typological frameworks, the vast number of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

Despite its important role for Fennoscandian archaeology, the applicability of dating by reference to shoreline displacement has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal here is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age in a relatively well sampled portion of Scandinavia, and in turn have this quantification inform the development of a formalised method for shoreline dating. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shoreline is inherently spatial. Taking inspiration from studies that have integrated various sources of spatio-temporal uncertainty through Monte Carlo simulation [e.g. @bevan2013a; @crema2012; @crema2010; @crema2015; @yubero-gomez2016], a similar approach is adopted here and adapted to the parameters of post-glacial sea-level change and the Stone Age settlement of southern Norway.

@@

Relative sea-level (RSL) can be defined as the mean elevation of the surface of the sea relative to land, or, more formally, the difference in elevation between the geoid and the surface of the Earth as measured from the Earth's centre [@shennan2015]. Variation in this relative distance follow from a range of effects [e.g. @milne2009]. Of central importance here is eustasy and isostasy. Eustatic sea-level is understood to be the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015]. The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans [@mörner1976]. Isostasy, on the other hand, pertains to adjustments in the crust to regain gravitational equilibrium relative to the underlying viscous mantle caused by mass loading and unloading, which occurs with glaciation and deglaciation. These effects causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight [@milne2015].

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g. @hughes2016; @stroeven2016, see Figure \@ref(fig:deglaciation)], the isostatic rebound has caused most areas of Norway to have been subjected to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g. @mörner1979; @svendsen1987]. In other words, the RSL has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is faster towards the centre of the ice sheet relative to the distal aspects. Thus, there is differential glacio-isostatic impact to a site’s location depending on it's relation to the ice sheet’s centre of mass, leading some areas on the outer coast to have had a more stable RSL or been subject to marine transgression [e.g. @romundset2015; @svendsen1987]. These conditions are directly reflected in the archaeological record. In areas where the sea-level has been stable over longer periods of time, people have often reused coastal site locations multiple times and over long time spans, creating a mix of settlement phases that are difficult to disentangle [e.g. @hagen1963; @reitan2009]. Transgression phases, on the other hand, can lead to complete destruction of the sites, bury them in marine sediments, or in the outermost periphery, submerge them [@bjerck2008; @glorstad2020]. Transgression can therefore lead to a hiatus in the archaeological record for certain sub-phases in the impacted areas despite the fact that there were likely coastal settlements during the inferred hiatuses. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 220m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @romundset2018; @sørensen1979]. This makes the region especially useful for evaluating the assumption of a shore-bound settlement pattern over a long and continuous time span.

```{r deglaciation, echo = FALSE, fig.cap = "Deglaciation at 1000 year intervals from c. 17--8 thousand years BCE. The study area defined later in the text is marked with a red outline (deglaciation data from Hughes et al. 2016, but see also Romundset et al. 2019 in relation to the study area).", out.width = "525px", fig.align = "center", regfloat = TRUE}

@@

The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than an earliest possible date for when a site could have been in use [see review by @nordqvist1999]. The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @glorstad2002; @glorstad2003; @glorstad2004; @jaksland2001; @jaksland2012; @jaksland2012a; @jaksland2014a; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below]. Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time. This is done by aggregating point estimates of shoreline dates in 100-, 200- or 500-year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @mjærum2022; @nielsen2021b; see also @jorgensen2020; @tallavaara2020]. In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963; @johansen1997; @mikkelsen1975a, 100].

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways [@brogger1905, 166; @bjerck1990; @bjerck2017; also discussed by @berg-hansen2009; @bergsvik2009]. The same logic has also been extended to the hinterland and inland regions, where sites are belived to be predominantly located along rivers and lakes [@brogger1905, 166; @glorstad2010, 57--87; but see also @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@prescott2020; cf. @solheim2021]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the development of settled farmsteads is to have led site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2020; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2012]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation and return to foraging in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227]. @nielsen2021b has recently argued that the initial uptake of agriculture in Early Neolithic south-eastern Norway is combined with a more complex settlement pattern, and that a simple foraging/agricultural dichotomy would underplay the variation present in the Early and Middle Neolithic settlement data [see also e.g. @amundsen2006; @solheim2012, 74; @ostmo1988]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line with the transition to the Early Neolithic around 3900 BCE, and finally a decisive shift with the Late Neolithic c. 2400 BCE.

Based on the generally accepted premise that most pre-Late Neolithic sites in south-eastern Norway located lower than the marine limit were situated on or close to the contemporaneous shoreline, it is common to err on the side of a shore-bound site location unless there is strong evidence to suggest otherwise. This is for example reflected in archaeological survey practices, which are often guided by both a digital and mental reconstruction of past sea-levels [e.g. @berg-hansen2009; @eskeland2017; @nummedal1923]. Similarly, following an excavation, if typological indicators in the assemblages correspond with available shoreline displacement curves, a shore-bound site location is often assumed, even if the typologically informed date span is too wide to decisively verify this. It is also common to combine this with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would appear to have been sensible if the site was not shore bound based on an archaeological understanding [e.g. @jaksland2014; @johansen1963; @nummedal1923]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present-day landscape it would make intuitive sense that the site was in use when the ocean reached closer to its elevation, as the site would have been accessible by means of watercraft. Although it appears that the arguments for such site locations are sensible and can for the most part be assumed to hold, comprehensive evaluations and attempts at quantification of this tendency are relatively few [see also @ilves2011].

One of the more extensive evaluations of the relationship between archaeological radiocarbon dates and RSL-change was done by Solheim and colleagues [@breivik2018; @solheim2020], who compared 102 radiocarbon dates from 33 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found an overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5% of the sites. However, where there was a discrepancy, the main occupation of the sites are still believed to have been shore-bound rather than associated with the deviating ^14^C-dates. This is based on typological and technological characteristics of the assemblages. Whether these mismatches represent later shorter visits that are responsible for the younger radiocarbon dates, or whether these dates are entirely erroneous can be difficult to evaluate [e.g. @persson2008; @schulke2020]. However, this distinction is not deemed critical here, as what is of interest is settlements and tool-production areas as evidenced by artefact inventories or multiple site features. Not remnants of stays as ephemeral to only be discernible by isolated features or dubious ^14^C-dates. The evaluation of the relevance of radiocarbon dates to settlement activity will here therefore be entirely dependent upon, and follow the discretion of the original excavation reports.

Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @akerlund1995; @åstveit2018; @solheim2020; see also @bjerck2008a; @kleppe1985; @ramstad2009]. This approach has a couple of limitations. First, the displacement curves are sometimes applied directly to larger study areas, analogous to what @borreggine2022 term a bathtub model, with only some studies having taken the variable uplift-rates into account when performing this comparison [e.g. @åstveit2018; @fossum2020; @møller1987; @persson2008]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that the conclusion supports the hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are actually the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation and help move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating.

# Data
To get at the relationship between sites and the contemporaneous shoreline, this analysis was dependent on identifying a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west. This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorenseninpress], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].

The shoreline displacement data used in this study is based on the so-called isolation basin method [e.g. @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments. Each basin thus represents a high precision sea-level index point (SLIP) which are combined using what has been termed the isobase method to devise a continuous time series for RSL-change adjusted to a common isobase. To minimise the impact of variable uplift rates, the cored basins are therefore located in as constrained of an area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more stark towards the north-east, which needs to be adjusted for in the case that any basins are located any significant distance from the common isobase that runs perpendicular to this uplift gradient (Figure \@ref(fig:map-iso)). The SLIPs indicate the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves, based on the present-day tidal range. For simplicity, the tidal range is assumed to have been the same throughout the Holocene [@sørensen2014, 44]. The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021 30cm at the standard port Helgeroa in Larvik]. Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs, and they are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is in part that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but they also contain potential error related to the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, to name but a few [e.g. @romundset2011; @romundset2019; for alternative approaches see e.g. @barnett2020; @cahill2016; @creel2022]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.
